---
title: "Predicting Airline Delay"
author: "Karen Chik"
date: "2/14/2017"
output: html_document
---


#### Read Data

Go to the airlines data site: 
http://stat-computing.org/dataexpo/2009/the-data.html. 
Read in the airlines data set for 2008 into a data frame.

```{r,echo=FALSE}
setwd("/Users/Karen/Desktop/Github/Flight Delays")
library(data.table)
air<-fread("2008.csv")
nrow(air)

```

#### Remove rows with missing data

```{r}
air<-na.omit(air)
nrow(air)
```

#### Regression

Fit one regression model each to explain "DepDelay" and "ArrDelay". Use a subset of 1 million rows of the data with no missing data. Keep the remaining data for out-of-sample testing. 

```{r}
#Setting relevant dummy variables
air$Jan <- ifelse(air$DayofMonth == 1,1,0)
air$Feb <- ifelse(air$DayofMonth == 2,1,0)
air$Mar <- ifelse(air$DayofMonth == 3,1,0)
air$Apr <- ifelse(air$DayofMonth == 4,1,0)
air$May <- ifelse(air$DayofMonth == 5,1,0)
air$Jun <- ifelse(air$DayofMonth == 6,1,0)
air$Jul <- ifelse(air$DayofMonth == 7,1,0)
air$Aug <- ifelse(air$DayofMonth == 8,1,0)
air$Sep <- ifelse(air$DayofMonth == 9,1,0)
air$Oct <- ifelse(air$DayofMonth == 10,1,0)
air$Nov <- ifelse(air$DayofMonth == 11,1,0)

air$Mon <- ifelse(air$DayOfWeek == 1,1,0)
air$Tues <- ifelse(air$DayOfWeek == 2,1,0)
air$Wed <- ifelse(air$DayOfWeek == 3,1,0)
air$Thurs <- ifelse(air$DayOfWeek == 4,1,0)
air$Fri <- ifelse(air$DayOfWeek == 5,1,0)
air$Sat <- ifelse(air$DayOfWeek == 6,1,0)

train<- sample(nrow(air),1000000)
test<-air[train, ]

Dep_Delay1<-lm(DepDelay~CRSDepTime+Distance+CarrierDelay+TaxiOut+WeatherDelay+NASDelay+SecurityDelay+LateAircraftDelay+Mon+Tues+Wed+Thurs+Fri+Sat+Jan+Feb+Mar+Apr+May+Jun+Jul+Aug+Sep+Oct+Nov+factor(UniqueCarrier),data=test)
summary(Dep_Delay1)

Arr_Delay1<-lm(ArrDelay~DepDelay+Distance+TaxiIn+CarrierDelay+WeatherDelay+SecurityDelay+LateAircraftDelay+Mon+Tues+Wed+Thurs+Fri+Sat+Jan+Feb+Mar+Apr+May+Jun+Jul+Aug+Sep+Oct+Nov+factor(UniqueCarrier),data=test)
summary(Arr_Delay1)

```


#### Take the fitted regression and predict delays using the remaining data from the no-missing data set. Report the absolute mean error in prediction. 

```{r}
#holdout sample

hold<-air[-train,]
library(lmtest)
Dep_Delay_hold<-lm(DepDelay~CRSDepTime+Distance+CarrierDelay+TaxiOut+WeatherDelay+NASDelay+SecurityDelay+LateAircraftDelay+Mon+Tues+Wed+Thurs+Fri+Sat+Jan+Feb+Mar+Apr+May+Jun+Jul+Aug+Sep+Oct+Nov+factor(UniqueCarrier),data=hold)
Dep_pred = predict(Dep_Delay_hold, data=hold) # Generate predicted probabilities
Actual<- hold$DepDelay
error<- Actual - Dep_pred
mean(abs(error))

Arr_Delay_hold<-lm(ArrDelay~DepDelay+Distance+TaxiIn+CarrierDelay+WeatherDelay+SecurityDelay+LateAircraftDelay+Mon+Tues+Wed+Thurs+Fri+Sat+Jan+Feb+Mar+Apr+May+Jun+Jul+Aug+Sep+Oct+Nov+factor(UniqueCarrier),data=hold)
Arr_pred = predict(Arr_Delay_hold, data=hold)
Actual<- hold$ArrDelay
error1<- Actual - Arr_pred
mean(abs(error1))
```
